HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-06 submission.Rmd
.

Prior to more involved feature engeneering steps, we transform the coordinate system by redefining the x-variable as the x-distance to the endzone (such that all play directions are from right to left and the relevant endzone is at zero), and changing the direction variable, such that zero degrees represents heading straight towards the corresponding endzone. As mentioned in the main text, we use player features from 5 defensive and 5 offensive player. First, we standardized their $x$ and $y$ coordinates with respect to the football and order the players according to their $y$ coordinates. That is the the first defender in our dataset is always the leftmost defensive player, while the first offensive player is always the rightmost one (offensive play direction is from left to right). Furthermore, for each player we compute distances to the football and their orientation with respect to the quarterback.

@@

As frequently mentioned, we face the problem of having only a small amount of relevant data available. In order to avoid overfitting problems, we therefore focus on a basic features set of 32 variables for our main results: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. However, using more thoroughly crafted features such as distances and orientation as described above, we can enlarge the feature set to 67 total variables: 30 distance variables (for each of the 10 relevant players the total distance, $x$-distances, and $y$-distances to the football), and 5 orientation variables (for each defender the orientation with respect to the QB).

@@

A hidden Markov model consists of an observed time series $\{y_t\}_{t=1}^T$ and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$. In this case, at every time point $t$, $y_t$ is the y-coordinate of the defensive player and $g_t$ proxies the offensives player to be guarded, i.e. the guarding allocation. 
The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}^{(1)} = \bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij})$, with $\gamma_{ij} = \Pr(g_t = j \mid g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $y_t$ are fully determined by the state that is currently active. More formally,
\begin{equation*}
f(y_t \mid g_1, \ldots, g_T, y_1, \ldots, y_{t-1},y_{t+1},\ldots,y_T) = f(y_t \mid g_t = j), \qquad j \in \{1, \ldots, N\},
\end{equation*}
which we denote by $f_j(y_t)$ in short.
In general, $f_j$ can be any density or probability mass function depending on the type of data and a typical choice is a parametric distribution with separate parameters for each latent state. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution with a mean that is fully determined by the current $y$-coordinate of defender $j \in \{1, \dots, N\}$ and a standard deviation that is fixed across all states but estimated from the data.

To fit the model, we use direct numerical likelihood maximization. The HMM likelihood for the motion of a specific defender in a specific play can be calculated based on the so-called *forward algorithm*. It effectively performs a summation over all possible latent state sequences in an efficient manner, rendering the computational complexity linear in the number of observations. Time series of different defenders within the same play, and of different plays are treated as independent, hence their likelihood contributions are summed to obtain the full likelihood of the training data. For practical implementation, we wrote a custom likelihood function in R, using the function `forward()` and other convenience functions from the R package `LaMa` (Koslik, 2024) to speed up computations. Furthermore, we used the R package `RTMB` to make this likelihood function compatible with automatic differentiation, making the numerical optimization process more efficient and robust. 
The parameters to be estimated are only the transition probability matrix $\boldsymbol{\Gamma}$ and the standard deviation of the Gaussian distribution. 
The initial distribution of each defender for each play comes from a deterministic assignment approach based on spatial proximity of the initial y-coordinates at the beginning of the time-series, i.e. the moment the line is set. Specifically, for each defender in a predefined set of potential defenders, we calculated their Euclidean distances to the five offensive players For each offense player, ranked in descending order by proximity, we assigned the closest defender to that player. The initial distribution was then set as a categorical distribution, where each defender was assigned a probability of 1 for their corresponding attacker and 0 for all others.

Having fitted an HMM to the data, we can use the model to predict the underlying state sequence based on the observations. This process is called state decoding and two main approaches exist. So-called *local decoding* constructs the conditional distributions
$$
\Pr(g_t = j \mid y_1, \dots, y_T)
$$
while *global decoding* using the *Viterbi-algorithm* finds the state sequence that maximizes the joint probability of the state sequence given the observations. Local decoding retains more probabilistic information as it provides a categorical state distribution for each time point while global decoding is more suitable for visualization purposes as it provides a single state sequence that is most likely to have generated the observations.
To obtain both the local state probabilities and the global state sequence, we used the functions `stateprobs()` and `viterbi()` that are also contained in the R package `LaMa`.

>>>>

Prior to more involved feature engeneering steps, we transform the coordinate system by redefining the x-variable as the x-distance to the endzone (such that all play directions are from right to left and the relevant endzone is at zero), and changing the direction variable, such that zero degrees represents heading straight towards the corresponding endzone. As mentioned in the main text, we use player features from 5 defensive and 5 offensive players. First, we standardized their $x$- and $y$-coordinates with respect to the football and ordered the players according to their $y$-coordinates, i.e., the the first defender in our dataset is always the leftmost defensive player, while the first offensive player is always the rightmost one (offensive play direction is from left to right). Furthermore, for each player we compute distances to the football and their orientation with respect to the quarterback.

@@

As frequently mentioned, we face the problem of only having a small amount of relevant data available. In order to avoid overfitting problems, we therefore focus on a basic feature set of 32 variables for our main results: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. However, using more thoroughly crafted features such as distances and orientation as described above, we can enlarge the feature set to 67 total variables: 30 distance variables (for each of the 10 relevant players the total distance, $x$-distances, and $y$-distances to the football), and 5 orientation variables (for each defender the orientation with respect to the QB).

@@

A hidden Markov model consists of an observed time series $\{y_t\}_{t=1}^T$ and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$. In this case, at every time point $t$, $y_t$ is the y-coordinate of the defensive player and $g_t$ proxies the offensives player to be guarded, i.e. the guarding assignment. 
The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}^{(1)} = \bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij})$, with $\gamma_{ij} = \Pr(g_t = j \mid g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $y_t$ are fully determined by the state that is currently active. More formally,
\begin{equation*}
f(y_t \mid g_1, \ldots, g_T, y_1, \ldots, y_{t-1},y_{t+1},\ldots,y_T) = f(y_t \mid g_t = j), \qquad j \in \{1, \ldots, N\},
\end{equation*}
which we denote by $f_j(y_t)$ in short.
In general, $f_j$ can be any density or probability mass function depending on the type of data and a typical choice is a parametric distribution with separate parameters for each latent state. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution with a mean that is fully determined by the current $y$-coordinate of defender $j \in \{1, \dots, N\}$ and a standard deviation that is fixed across all states but estimated from the data.

To fit the model, we use direct numerical likelihood maximization. The HMM likelihood for the motion of a specific defender in a specific play can be calculated based on the so-called *forward algorithm*. It effectively performs a summation over all possible latent state sequences in an efficient manner, rendering the computational complexity linear in the number of observations. Time series of different defenders within the same play and of different plays are treated as independent, hence their likelihood contributions are summed to obtain the full likelihood of the training data. For practical implementation, we wrote a custom likelihood function in R, using the function `forward()` and other convenience functions from the R package `LaMa` (Koslik, 2024) to speed up computations. Furthermore, we used the R package `RTMB` to make this likelihood function compatible with automatic differentiation, making the numerical optimization process more efficient and robust. 
The parameters to be estimated are only the transition probability matrix $\boldsymbol{\Gamma}$ and the standard deviation of the Gaussian distribution. 
The initial distribution of the guarding assignment for each defender in each play arises from a deterministic assignment approach, based on spatial proximity at the beginning of the time-series, i.e., the moment the line is set. Specifically, for each defender in the predefined set of defenders considered, we compute a weighted Euclidean distance (giving more weight to the $y$-coordinate) to the five offensive players that could potentially be guarded. The initial distribution is then set to 1 for the closest defender under this metric and 0 otherwise. While this approach proved to be sufficiently reliable in this application, it may be finetuned in future iterations.

Having fitted an HMM to the data, we can use the model to predict the underlying state sequence based on the observations. This process is called *state decoding* and two main approaches exist. So-called *local decoding* constructs the conditional distributions
$$
\Pr(g_t = j \mid y_1, \dots, y_T)
$$
while *global decoding* using the *Viterbi-algorithm* finds the state sequence that maximizes the joint probability of the state sequence given the observations. Local decoding retains more probabilistic information as it provides a categorical state distribution for each time point, while global decoding is more suitable for visualization purposes as it provides a single state sequence that is most likely to have generated the observations.
To obtain both the local state probabilities and the global state sequence, we used the functions `stateprobs()` and `viterbi()` that are also contained in the R package `LaMa`.
